Leader functions and qualities across 59 cultures: an exploratory analysis

### 2019-11-27 leadership_across_cultures_contexts.Rmd
Working on introduction. Additions to bib file.

---
title: "Leader functions and qualities across 59 cultures: an exploratory analysis"
author: "Zachary H. Garfield, Kristen L. Syme, and Edward H. Hagen"
date: "4/25/2019"
output: 
  bookdown::html_document2
bibliography: Leader2.bib
---

@@

Across human societies, the influence of leaders stands to substantially shape the nature of group structure and group outcomes. Simultaneously, social norms, pressures from competing leaders or rival or cooperative groups, and followers preferences can greatly constrain a leader's agenda. The interactions between the strategies of individual leaders, pressures of followership, and cultural forces and group dynamics likely play an out-sized role in contributing to organizational variation across human populations including social and economic trajectories of populations [@wiessner_risk_1982], and cultural evolutionary processes [@henrich_big_2015]. Leadership and followership are critical in understanding group structure and have been the focus of much scientific attention across psychology, sociology, political science, and managerial studies. Until recently, this very large body of work was relatively disconnected from approaches to leadership in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010 and @kaplan_evolutionary_2009], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @cheng_two_2013; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016]. This approach as been informative. Drawing on the study of leadership in evolutionary perspective and across diverse, small-scale and traditional societies can help develop more generalizable theories and offer applications to shape organizational structure in ways which are more consistent with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019]. Nonetheless, much of this work has focused on community or political leaders, often overlooking the nature of leadership and followership across social domains. Moreover, while the benefits leaders accrue for their services have received substantial attention the costs of leadership for both leaders and followers have also been relatively under-investigated. Current outstanding questions for evolutionary leaderhip scholars include the degree to which the nature of leadership is shaped by population-level variation and the costs and benefits leaders and followers encur across distinct leadership systems.

@@

We first briefly review current theoretical, ethnographic, and empirical perspectives on the qualities of leaders, the functions they serve, and the costs and benefits of leadership for both leaders and followers across human societies, including WEIRD and non-WEIRD populations. We then outline the aims and methods of the current cross-cultural study, before presenting results and interpretations.

@@

Leadership as a phenomena is a universal feature of human groups; among all societies there are particular contexts in which one individual maintains greater influence. Also, adopting the broad definition of leadership given above, parents can be defined as leaders of nuclear families. Strong community leadership, however, is common cross-culturally [@garfield_review_2019; @lewis_leaders_1974].

The emergence of leadership and the functions leaders serve across cultures is closely tied to social and political structure.

Among ethnographically described small-scale societies which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and tends to emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015]. Leadership tends to be context-specific and leaders in such contexts often draw on persuasive capacities and leverage community-based respect, their greater knowledge and social support to express their influence. The theoretical and ethnographic literature suggests these demands most commonly arise in the contexts of conflict resolution and collective actions. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers. Representing the group and managing both internal and external relationships and conflicts are expected to be the most salient functions of community leaders.

Within communities, leaders may emerge whenever groups of individuals form. The role of leaders in ritual contexts, such as in initiations, and in providing social services, including facilitating marriages and organizing social activities, are also relatively common.

Increased sedentarization and defensibility of resources is often associated with both social stratification and an institutionalized managerial role of community leaders. Among horticulturalists and agriculturalists managerial leadership can promote for more efficient use of shared resources such as water reserves and cooperative labor. Pastoralists and coastal populations can face similar pressures surrounding grazing lands and fishing access. In the wake of economic intensification, the function of leadership becomes increasing associated with the distribution of resources.

Defensibility of resources is also often associated with increased territoriality and increased pressures of inter-group conflicts. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territory are often forced to manage military forces and develop offensive and defensive strategies in warfare, more so than mobile populations.

In WEIRD societies, community leaders are primarily heads of governmental and corporate bodies with specific responsibilities and functions associated with their position. The structure of these organizations are generally organized hierarchically with nested bodies and a formal internal 'chain of command.' At the higher levels of large-scale organizations, leaders primarily serve as representatives in inter-group interactions, facilitate the creation and adoption of policies, and serve as a figure head embodying the strategic vision of the organization. They are not often involved in resolving immediate conflicts, such as disputes between individuals or nested bodies, however, they are often the final end-point in a decision making hierarchy and the results of a single decision of these leaders can trickle down through industry or government with global ramifications.

@@

Despite immense variation, there do seem to be some commonalities among the qualities of community leaders. Across human societies, when strong community leaders exist, they tend to be high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have differential influence. Elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have increased influence over community decisions across cultures, including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations.

Among small-scale societies with relative egalitarianism and local political autonomy community leaders, in addition to being older, respected males, tend to be individuals with large social networks and strong social support. Physically, leaders across multiple contexts tend to be taller and more physically formidable. Such physical embodiment might equip individuals for more effective leadership in the context of inter-group conflicts as well as in the context of intra-group dispute settlement and in management of collective actions. Charisma and oratory skills are common properties of leaders, but are especially important in the context of egalitarianism where individual leaders must rely on persuasion and more subtle forms of influence to convince community members to come to agreement.

@@

The ability to shape group dynamics can provide enormous individual benefits to leaders, even if 'just for a while.' Fundamentally, leaders in any context are able to promote group policies which either align with their individual interests or which are not especially costly for themselves.

@@

Institutionalized leadership positions are also typically associated with increased social status.

@@

The unique costs associated with leadership positions have received less attention, relative to the benefits of leadership.

@@

Evolutionary theoretical models and empirical studies often seek to integrate findings within a broad framework of human cultural diversity. The universality of human leadership systems and the incredible diversity among the qualities of leaders across cultures and context and the functions they serve is difficult to capture. Scholars are then left to haphazardly self-select ethnographic cases or rely on the self-selected reviews of others. Here, we provide the first broad and near comprehensive view of the qualities and functions of leaders across from a representative cross-cultural sample. We also aim to understand how leader qualities relate to their functions and how these relationships vary across social contexts and levels of cultural complexity. We hope these empirical results will facilitate more fine-grained approaches within leadership studies and greater theoretical and empirical work beyond community leadership, while also revealing the importance of a broad, cross-cultural perspective in understanding the psychology and cultural patterning of perhaps the most foundation component of human social behavior and social organization.

@@

The operationalized qualities and functions (See Table X) were developed from the ethnographic text records in the database. First ZG read a subset of text records and developed initial variable lists and operationalizations of the qualities and functions in those text records. Then, KS coded all 1212 text records (in a random order) for the operationalized qualities, functions, cost, and benefits, creating new variables and operationalizations for additional qualities and functions encountered. After coding half of the random sample KS and ZG revised the operational definitions to produce the final list of variables in Tables X. ZG then followed the same process of coding the entire sample of 1212 text records for the X operationalized variables. DISCUSS IRR. ZG and KS then discussed every instance of disagreement between their respective coding to produce a consensus matrix, which is used in the analyses here and added to the published data repository.

Coding operationalizations of the costs and benefits of leadership were developed *a priori* based on our reading of the evolutionary theoretical literature from anthropology and biology. For both followers and leaders, we code if there is evidence of costs or benefits in the following categories: social status/reputation, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, other resources, and other costs or benefits not captured by these categories. See Table X.

@@

## Distribution of text records by group type and sex

The type of group to which a leader text referred was coded into seven categories. Figure \@ref(fig:groupsex) indicates the proportion of text records that were categorized into each of these groups; it also indicates the proportion of male and female leaders of each type of group.

```{r groupsex, fig.width=15, fig.cap="The distribution of text records by group type and leader sex. Areas of the bars are proportional to the number of text records in that category. There were no female leaders in any military or religious group."}
plot_group_sex
```

@@

## Benefits and costs of leadership and followership

>>>>

---
title: "Leader functions and qualities across 59 cultures: an exploratory analysis"
author: "Zachary J. Garfield, Kristen L. Syme, and Edward H. Hagen"
date: "4/25/2019"
output: 
  bookdown::html_document2
bibliography: Leader2.bib
---

@@

Across human societies, the influence of leaders stands to substantially shape the nature of group structure and group outcomes. Simultaneously, social norms, pressures from competing leaders or rival or cooperative groups, and followers preferences can greatly constrain a leader's agenda. The interactions between the strategies of individual leaders, pressures of followership, and group dynamics are likely to strongly contribute to organizational variation across human populations including social and economic trajectories [@wiessner_risk_1982] and cultural evolutionary processes [@henrich_big_2015].

Leadership and followership are critical in understanding group structure and have been the focus of much scientific attention across psychology, sociology, political science, and managerial studies. Until recently, this very large body of work was relatively disconnected from approaches to leadership in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010 and @kaplan_evolutionary_2009; @von_rueden_roots_2014], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @cheng_two_2013; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016]. This integration as been informative. Drawing on the study of leadership in evolutionary perspective and across diverse, small-scale and traditional societies can help develop more generalizable theories and offer applications to shape organizational structure in ways which are more consistent with human evolutionary history and evolved leader-follower psychologies [@von_rueden_leadership_2015; @garfield_review_2019]. Nonetheless, much of this work has focused on community or political leaders, often overlooking the nature of leadership and followership across social domains and at all levels of social organization. Moreover, while the benefits leaders accrue for their services have received substantial attention the costs of leadership for both leaders and followers have also been relatively under-investigated. Current outstanding questions for evolutionary leaderhip scholars include the degree to which the nature of leadership is shaped by population-level variation and the costs and benefits leaders and followers encur across distinct leadership systems.

@@

We first briefly review current theoretical, ethnographic, and empirical perspectives on the qualities of leaders, the functions they serve, and the costs and benefits of leadership for both leaders and followers across human societies, including WEIRD and non-WEIRD populations. We then outline the aims and methods of the current cross-cultural study, before presenting results and interpretations. Through novel empirical and exploratory analyses encompassing the full range of cultural diversity and social contexts, we aim to advance leadership studies across disciplines and provide researchers with a foundation for more detailed approaches.

@@

Leadership is a universal feature of human groups; among all societies there are particular contexts in which one individual maintains greater influence over group behavior and outcomes. Additionally, under the broad definition of leadership given above, parents can be defined as leaders of nuclear families. Strong community leadership, however, is common cross-culturally including among hunter-gatherers and relatively small-scale societies [@garfield_review_2019; @lewis_leaders_1974].

The nature of leaderhsip across human societies and social contexts is highly diverse and the emergence of leaders and the functions they serve are closely tied to social and political structure. Among ethnographically described small-scale societies which generally lack multi-level, institutionalized social and political structures, leadership at the community level is typically informal and ephemeral and tends to emerge contextually when demands of coordination or conflict resolution require a managerial presence [@von_rueden_leadership_2015]. Leadership tends to be context-specific and leaders in such contexts often draw on persuasive capacities and leverage community-based respect and greater knowledge and social support to achieve or maintain community influence. The theoretical and ethnographic literature suggests these demands most commonly arise in the contexts of conflict resolution and collective actions. Demands of mediation in within-group conflict resolution, although variable in scale, are probably a universal feature of human groups. Demands of between-group interactions, including both conflict and cooperation are also very common across small-scale societies, including hunter-gatherers. Representing the group and managing both internal and external relationships and conflicts are expected to be the most salient functions of community leaders.

Within communities, leaders may emerge whenever groups of individuals form. The role of leaders in ritual contexts, such as in initiations, and in providing social services, including facilitating marriages and organizing social activities, are also expected to be common functions leaders serve.

Increased sedentarization and defensibility of resources is often associated with both social stratification and an institutionalized managerial role and stronger authority of community leaders. Among horticulturalists and agriculturalists managerial leadership can promote for more efficient use of shared resources such as water reserves and cooperative labor. Pastoralists and coastal populations often face similar pressures surrounding grazing lands and fishing access. In the wake of economic intensification, the function of leadership becomes increasing associated with the distribution of resources.

Defensibility of resources necessarily creates increased oppurtunities for resource management and is often associated with increased territoriality and increased pressures of inter-group conflicts. Community leaders among populations more reliant on domesticated plant foods, livestock, and specific territories are often required to manage military forces and develop both offensive and defensive strategies in warfare, more so than mobile populations that rely on disperrsed, less defesnible resources.

In WEIRD societies, community leaders are primarily heads of governmental and corporate bodies with institutionalized responsibilities and functions associated with their position. The structure of these organizations are generally organized hierarchically with nested bodies and a formal internal 'chain of command.' At the higher levels of large-scale organizations, leaders primarily serve as representatives in inter-group interactions, facilitate the creation and adoption of policies, and serve as figure heads embodying the strategic vision of the organization. They are not often involved in resolving immediate conflicts, such as disputes between individuals or nested bodies, however, they are often the final end-point in a decision making hierarchy and the results of a single decision of these leaders can trickle down through industry or government with global ramifications.

@@

Despite immense variation, there do seem to be some commonalities among the qualities of community leaders. Across human societies, when strong community leaders emerge, they tend to be high status males who embody the ideal qualities most respected by the group and necessary for successful leadership in the domains in which they have differential influence. Elders, respected individuals, and highly intelligent or knowledgeable individuals tend to have increased influence over community decisions across cultures, including among the most egalitarian hunter-gatherers, stratified chiefdoms, and contemporary state-level organizations.

Among small-scale societies with relative egalitarianism and local political autonomy community leaders, in addition to being older, respected males, tend to be individuals with large social networks and strong social support. Physically, leaders across multiple contexts tend to be taller and more physically formidable. Such embodied capital might equip individuals for more effective leadership in the context of inter-group conflicts as well as in the context of intra-group dispute settlement and in management of collective actions. Charisma and oratory skills are common properties of leaders, but are especially important in the context of egalitarianism where individual leaders must rely on persuasion and more subtle forms of influence to convince community members to come to agreement.

@@

The ability to shape group dynamics can provide enormous individual benefits to leaders, even if 'just for a while.' Fundamentally, leaders in any context have the oppurtunity to promote group-level policies which either align with their individual interests or which are not especially costly for themselves, their kin, or their close social partners.

@@

Institutionalized leadership positions are also often, if not universally, associated with increased social status. The causal relationship between high social status and community or political leadership is complex and likely to be context-specific and bi-directional.

@@

The unique costs associated with leadership positions have received less attention, relative to the benefits of leadership across disciplines.

@@

Evolutionary theoretical models and empirical studies often seek to integrate findings within a broad framework of human cultural diversity. The universality of human leadership systems and the incredible diversity among the qualities of leaders across cultures and contexts and the functions they serve is difficult to capture. Scholars are then left to haphazardly self-select ethnographic cases or rely on the self-selected reviews of others for generalizations. Here, we provide the first broad and near comprehensive view of the qualities and functions of leaders across human societies from a representative cross-cultural sample. We also provide the first comparative empirical view describing the costs and benefits of leadership for both leaders nad followers in an evolutionary biological framework. We also aim to understand how the qualities and functions of leaders vary across social contexts and levels of cultural complexity. We hope these exploratory results will facilitate more fine-grained approaches within leadership studies and greater theoretical and empirical work beyond community leadership, while also revealing the importance of a broad, cross-cultural perspective in understanding the psychology and cultural patterning of perhaps the most foundational component of human social behavior and social organization.

@@

The operationalized qualities and functions (See Table X) were developed from the ethnographic text records in the database. First ZG read a subset of text records and developed initial variable lists and operationalizations of the qualities and functions in those text records. Then, KS coded all 1212 text records (in a random order) for the operationalized qualities, functions, cost, and benefits, while also creating new variables and operationalizations for additional qualities and functions encountered. After coding half of the random sample KS and ZG revised the operational definitions to produce the final list of variables in Tables X. ZG then followed the same process of coding the entire sample of 1212 text records for the X operationalized variables. DISCUSS IRR. ZG and KS then discussed every instance of disagreement between their respective coding to produce a consensus matrix, which is used in the analyses here and added to the published data repository (a dataframe named leader_text2).

Coding operationalizations of the costs and benefits of leadership were developed *a priori* based on our reading of the evolutionary theoretical literature from anthropology and biology. For both followers and leaders, we code if there is evidence of costs or benefits in the following categories: social status, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, material resources, and other miscelaneous costs or benefits not captured by these categories. See Table X.

@@

@@
